---
title: "Numerical Data Applications Solutions"
author: "Professor Bradley Warner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE,results='hide',message=FALSE}
library(openintro)
library(mosaic)
library(tidyverse)
```


# Exercises

Create this Rmd file for the work including headers, file creation data, and explanation of your work. Make sure your plots have a title and the axes are labeled. We are asking you to do more in this application to get ready for your Oral Board.

1. **Mammals exploratory**  

Data were collected on 39 species of mammals distributed over 13 orders.  The data is in the `openintro` package as `mammals`

a. Using help, report the units for the variable `BrainWt`.  
```{r eval=FALSE}
?mammals
```


b. Using `inspect` how many variables are numeric?  

```{r}
inspect(mammals)
```

c. What type of variable is `danger`?

Categorical

d. Create a density plot of `total_sleep` and describe the distribution.  

```{r}
gf_dens(~total_sleep,data=mammals)
```


e. Create a boxplot of `life_span` and describe the distribution.  

```{r}
gf_boxplot(~life_span,data=mammals)
```


f. Report the mean and median life span of a mammal.  

```{r}
mean(~life_span,data=mammals,na.rm=TRUE)
```

```{r}
median(~life_span,data=mammals,na.rm=TRUE)
```

g. Calculate the summary statistics for `LifeSpan` broken down by `Danger`.

```{r}
favstats(life_span~danger,data=mammals)
```



2. **Mammals life spans**  

Continue using the `mammals` data set.

a. Create side-by-side boxplots for `life_span` broken down by `exposure`. Note: you will have to change `exposure` to a `factor()`. Report on any findings.  

```{r}
mammals %>%
gf_boxplot(life_span~factor(exposure))
```



b. What happened to the median and third quartile in exposure group 4?

```{r}
favstats(life_span~factor(exposure),data=mammals)
```

c. Create overlapping density plots. What are the shorcomings of this plot?

```{r}
gf_dens(~life_span,color=~factor(exposure),data=mammals)
```

d. Create a new variable `exposed` that is a factor with level `Low` if exposure is `1` or `2` and `High` otherwise.

```{r}
mammals <- mammals %>%
  mutate(exposed=factor(ifelse((exposure==1)|(exposure==2),"Low","High")))
```


```{r}
inspect(mammals)
```


e. Repeat part c with the new variable.

```{r}
gf_dens(~life_span,color=~exposed,data=mammals)
```

3. **Mammals life spans continued**  

a. Create a scatterplot of life span versus length of gestation.  

```{r}
mammals %>%
gf_point(life_span~gestation)
```


b. What type of an association is apparent between life span and length of gestation?   

It is a weak positive association.

c. What type of an association would you expect to see if the axes of the plot were reversed, i.e. if we plotted length of gestation versus life span?  

The same as this is observational data there is no reason to beliee is a causal relationship just by looking at the data. Switching the axis will preserve the association.

d. Create the new scatterplot suggested in c.  

```{r}
mammals %>%
gf_point(gestation~life_span)
```

e. Are life span and length of gestation independent? Explain your reasoning. 

No there is an association and it appears to be linear. If the plot looked like a "shotgun" blast, we would consider the variables to be independent. However, remember there may be confounding variables that could impact the association between these variables.

