---
title: "Additional Hypothesis Test Applications Solutions"
author:
- Lt Col Ken Horton
- Lt Col Kris Pruitt
- Professor Bradley Warner
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{multirow}
   - \usepackage{multicol}
   - \usepackage{changepage}
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align='center')
knitr::opts_chunk$set(out.width = "75%")
library(knitr)
library(mosaic)
library(tidyverse)
```

\newcommand{\E}{\mbox{E}}
\newcommand{\Var}{\mbox{Var}}
\newcommand{\Cov}{\mbox{Cov}}
\newcommand{\Prob}{\mbox{P}}
\newcommand*\diff{\mathop{}\!\mathrm{d}}

\indent 1. HFI   

Choose another freedom variable and a variable you think would strongly correlate with it. The `openintro` package contains the data set `hfi`. Type `?openintro::hfi` in the Console window in RStudio to learn more about the variables. 
    

```{r warning=FALSE,message=FALSE}
hfi<-tibble(read_csv("data/hfi.csv"))
```    
    
a. Produce a scatterplot of the two variables.  

We selected `pf_expression_influence` as it is a measure of laws and regulations that influence media content. We kept `pf_score` because it is a measure of personal freedom in a country. Our thought is these should be correlated.


```{r warning=FALSE}
gf_lm(pf_score~pf_expression_influence,data=hfi,color="black") %>%
  gf_theme(theme_bw()) %>%
  gf_point(alpha=0.3) %>%
  gf_labs(title="Personal freedom score versus Control on media",
          x="Laws and regulations that influence media content",
          y="Personal freedom score")
```


b. Quantify the strength of the relationship with the correlation coefficient.

```{r}
hfi %>%
  summarise(cor(pf_expression_influence, pf_score, use = "complete.obs"))
```

c. Fit a linear model. At a glance, does there seem to be a linear relationship?

```{r m1}
m2 <- lm(pf_score ~ pf_expression_influence, data = hfi)
```


```{r}
summary(m2)
```


d. How does this relationship compare to the relationship between 
    `pf_expression_control` and `pf_score`? Use the $R^2$ values from the two 
    model summaries to compare. Does your independent variable seem to predict
    your dependent one better? Why or why not?
    
    The adjuste $R^2$ is a little smaller so the fit is not as good.
    
e. Display the model diagnostics for the regression model analyzing this relationship.


**Linearity**: 

```{r residuals}
ggplot(data = m2, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x="Fitted values",y="Residuals",title="Residual analysis") +
  theme_bw()
```

There does appear to be some type of fluctuation but so the linear model may not be appropriate. 

**Nearly normal residuals**: 

```{r}
ggplot(data = m2, aes(x = .resid)) +
  geom_histogram(binwidth = .4) +
  xlab("Residuals") +
  theme_bw()
```

or a normal probability plot of the residuals.

```{r}
ggplot(data = m2, aes(sample = .resid)) +
  stat_qq() +
  theme_bw() +
  geom_abline(slope=1,intercept = 0)
```

No, the sample is small but it appears the residual are skewed to the left.

**Constant variability**:

> **Exercise**: 
Based on the residuals vs. fitted plot, does the constant variability condition 
    appear to be met?

Yes, the width of the plot seems constant with the exception of some extreme points.



f. Predict the response from your explanatory variable for a value between the median and third quartile. Is this an overestimate or an underestimate, and by how much? 

```{r}
summary(hfi$pf_expression_influence)
```


```{r}
predict(m2,newdata=data.frame(pf_expression_influence=6))
```
We thus predict a value of 7.53 for the `pf_score`.

The observed value is 7.96, an average of 42 data points. We tend to underestimate the observed value.


```{r}
library(broom)
```

```{r}
augment(m2) %>%
   filter(pf_expression_influence==6) %>%
    summarize(ave=mean(pf_score),n=n())
```




## File Creation Information 

  * File creation date: `r Sys.Date()`
  * Windows version: `r win.version()`
  * `r R.version.string`
  * `mosaic` package version: `r packageVersion("mosaic")`
  * `tidyverse` package version: `r packageVersion("tidyverse")`
  * `openintro` package version: `r packageVersion("openintro")`
  * `broom` package version: `r packageVersion("broom")`
 




