---
title: "Additional Hypothesis Tests Notes"
author:
- Lt Col Ken Horton
- Lt Col Kris Pruitt
- Professor Bradley Warner
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{multirow}
   - \usepackage{multicol}
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align='center')
knitr::opts_chunk$set(out.width = "75%")
library(knitr)
library(mosaic)
library(tidyverse)
```

\newcommand{\E}{\mbox{E}}
\newcommand{\Var}{\mbox{Var}}
\newcommand{\Cov}{\mbox{Cov}}
\newcommand{\Prob}{\mbox{P}}
\newcommand{\diff}{\,\mathrm{d}}

## Objectives

1) Using `R`, generate a linear regression model and use it to produce a prediction model.

2) Using plots, check the assumptions of a linear regression model.

## Introduction 

The Human Freedom Index is a report that attempts to summarize the idea of "freedom"
through a bunch of different variables for many countries around the globe.
It serves as a rough objective measure for the relationships between the different
types of freedom - whether it's political, religious, economical or personal freedom -
and other social and economic circumstances. The Human Freedom Index is an annually
co-published report by the Cato Institute, the Fraser Institute, and the Liberales
Institut at the Friedrich Naumann Foundation for Freedom.

In this case study, you'll be analyzing data from Human Freedom Index reports from 2008-2016.
Your aim will be to summarize a few of the relationships within the data both 
graphically and numerically in order to find which variables can help tell a story
about freedom. This will be done using the tool of regression.

Again, like our previous case studies, this lesson will introduce many of the ideas of the block. Don't worry if they seem difficult and you feel overwhelmed a bit, we will come back to the ideas in the following lessons.

### Load packages

Let's load the packages.

```{r message=FALSE,eval=FALSE}
library(tidyverse)
library(mosaic)
```



### The data and exploratory analysis

The data we're working with is in the file called `hfi.csv` under the `data` folder. The name `hfi` is 
short for Human Freedom Index. 

> **Exercise**  
Read the data into `R`. What are the dimensions of the dataset?

```{r warning=FALSE,message=FALSE}
hfi<-tibble(read_csv("data/hfi.csv"))
```

```{r}
dim(hfi)
```

There are 1458 observations and 123 variables in the data frame. 

>**Exercise**  
Create summaries of the first 10 variables in the data. We just don't want a large printout.

```{r warning=FALSE,message=FALSE}
inspect(hfi[,1:10])
```

>**Exercise**  
Create a scatter plot to display the relationship between the personal
    freedom score, `pf_score`, as the response and `pf_expression_control` as the predictor.
    Does the relationship look linear? If you knew a country's `pf_expression_control`,
    or its score out of 10, with 0 being the most, of political pressures and 
    controls on media content, would you be comfortable using a linear model to
    predict the personal freedom score?


```{r warning=FALSE}
gf_lm(pf_score~pf_expression_control,data=hfi,color="black") %>%
  gf_theme(theme_bw()) %>%
  gf_point(alpha=0.3) %>%
  gf_labs(title="Personal freedom score versus Expression control",
          x="Political pressures and controls on media",
          y="Personal freedom score")
```
```{r warning=FALSE}
ggplot(hfi,aes(x=pf_expression_control,y=pf_score))+
  geom_point(alpha=0.3) +
  theme_bw()+
  geom_lm(color="black")+
  labs(title="Personal freedom score versus Expression control",
          x="Political pressures and controls on media",
          y="Personal freedom score")
```
The relationship does look linear. We should be uncomfortable using the model at the end points. That is because there are less points at the edge and and linear estimation has larger variance at the endpoints, the predictions at the endpoints is more suspect. 

> **Exercise**  
The relationship looks linear, quantify the strength of the
relationship with the correlation coefficient.

```{r}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Note that we set the `use` argument to "complete.obs" since there are some observations with missing values, NA.


## Sum of squared residuals

In this section, you will use an interactive function to investigate what we mean by "sum 
of squared residuals". You will need to run this function in your console. Running the function also requires that the `hfi` data set is loaded in your environment, which we did above.


Think back to the way that we described the distribution of a single variable. 
Recall that we discussed characteristics such as center, spread, and shape. It's
also useful to be able to describe the relationship of two numerical variables, 
such as `pf_expression_control` and `pf_score` above.

> **Exercise**  
Looking at your scatterplot above, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.
    
We would say that there is a strong positive linear relationship between the two variables.

Just as you've used the mean and standard deviation to summarize a single variable, 
you can summarize the relationship between these two variables by finding the 
line that best follows their association. Use the following interactive 
function to select the line that you think does the best job of going through 
the cloud of points.

First we must remove missing values from the data set and to make the visualization easier, we will just randomly sample 30 of the data points. We included `hf_score` because we will need it later.

```{r}
set.seed(4011)
hfi_sub <- hfi %>%
  select(pf_expression_control,pf_score,hf_score) %>%
  drop_na() %>%
  group_by(pf_expression_control) %>%
  slice_sample(n=1)
```


We used the function `slice_sample()` to ensure we have unique values of `pf_expression_control` is our sample.

>**Exercise**  
In your `R` council, run the code above to create the object `hfi_sub`. You are going to have to load packages and read in the `hfi` data set. Then execute the next set of code. Pick two locations in the plot to create a line. Record the sum of squares.

```{r echo=FALSE}
# If you need a function to do this, here it is
plot_ss <- function (x, y, data, showSquares = FALSE, leastSquares = FALSE) 
{
    x <- eval(substitute(x), data)
    y <- eval(substitute(y), data)
    plot(y ~ x, asp = 1, pch = 16)
    if (leastSquares) {
        m1 <- lm(y ~ x)
        y.hat <- m1$fit
    }
    else {
        cat("Click two points to make a line.")
        pt1 <- locator(1)
        points(pt1$x, pt1$y, pch = 4)
        pt2 <- locator(1)
        points(pt2$x, pt2$y, pch = 4)
        pts <- data.frame(x = c(pt1$x, pt2$x), y = c(pt1$y, pt2$y))
        m1 <- lm(y ~ x, data = pts)
        y.hat <- predict(m1, newdata = data.frame(x))
    }
    r <- y - y.hat
    abline(m1)
    oSide <- x - r
    LLim <- par()$usr[1]
    RLim <- par()$usr[2]
    oSide[oSide < LLim | oSide > RLim] <- c(x + r)[oSide < LLim | 
        oSide > RLim]
    n <- length(y.hat)
    for (i in 1:n) {
        lines(rep(x[i], 2), c(y[i], y.hat[i]), lty = 2, col = "blue")
        if (showSquares) {
            lines(rep(oSide[i], 2), c(y[i], y.hat[i]), lty = 3, 
                col = "orange")
            lines(c(oSide[i], x[i]), rep(y.hat[i], 2), lty = 3, 
                col = "orange")
            lines(c(oSide[i], x[i]), rep(y[i], 2), lty = 3, col = "orange")
        }
    }
    SS <- round(sum(r^2), 3)
    cat("\r                                ")
    print(m1)
    cat("Sum of Squares: ", SS)
}
```


```{r plotss-expression-score}
plot_ss(x = pf_expression_control, y = pf_score, data = hfi_sub)
```

Once you've picked your two locations, the line you specified will be shown in
black and the residuals in blue. Residuals are the difference between the observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes
the sum of squared residuals. To visualize the squared residuals, you can rerun 
the plot command and add the argument `showSquares = TRUE`.

```{r plotss-expression-score-squares, eval=FALSE}
plot_ss(x = pf_expression_control, y = pf_score, data = hfi_sub, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and
intercept of your line as well as the sum of squares.

>**Exercise**:  
 Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?

## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the 
line that minimizes the sum of squared residuals, through trial and error. 
Instead, you can use the `lm()` function in `R` to fit the linear model (a.k.a. 
regression line).

```{r m1}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi_sub)
```

The first argument in the function `lm` is a formula that takes the form 
`y ~ x`. Here it can be read that we want to make a linear model of 
`pf_score` as a function of `pf_expression_control`. The second argument specifies
that `R` should look in the `hfi_sub` data frame to find the two variables. This should be
familiar to us since we have been doing this when we used the `mosaic` package.

The output of `lm` is an object that contains all of the information we need 
about the linear model that was just fit. We can access this information using 
the summary function.

```{r summary-m1}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe 
the model is shown at the top. After the formula you find the five-number 
summary of the residuals. The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `pf_expression_control`.
With this table, we can write down the least squares regression line for the 
linear model:

$$
  \hat{pf\_score} = 5.02721 + 0.41988 \times pf\_expression\_control
$$

At least these are the values we got on our machine using the *seed* provided. Yours may differ slightly. One last piece of information we will discuss from the summary output is the `Multiple R-squared`, or more simply, $R^2$. The $R^2$ value represents the proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 72.48% of the variability in `pf_score` is explained by `pr_expression_control`.

>**Exercise**:  
Fit a new model that uses `pf_expression_control` to predict `hf_score`, or
    the total human freedom score. Using the estimates from the `R` output, write 
    the equation of the regression line. What does the slope tell us in the 
    context of the relationship between human freedom and the amount of political
    pressure on media content?

```{r}
m2<-lm(hf_score ~ pf_expression_control, data = hfi_sub)
```

```{r}
summary(m2)
```

$$
  \hat{hf\_score} = 5.45660 + 0.30707 \times pf\_expression\_control
$$
As the political pressure increases by one unit, the **average** human freedom score increases by 0.307.

## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on top.

```{r reg-with-line,warning=FALSE,message=FALSE}
ggplot(data = hfi_sub, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE,color="black") +
  theme_bw()+
  labs(title="Personal freedom score versus Expression control",
          x="Political pressures and controls on media",
          y="Personal freedom score")
```

Here, we are literally adding a layer on top of our plot. `geom_smooth` creates
the line by fitting a linear model. It can also show us the standard error `se`
associated with our line, but we'll suppress that for now.

This line can be used to predict $y$ at any value of $x$. When predictions are 
made for values of $x$ that are beyond the range of the observed data, it is 
referred to as *extrapolation* and is not usually recommended. However, 
predictions made within the range of the data are more reliable. They're also 
used to compute the residuals.

>**Exercise**:  
If someone saw the least squares regression line and not the actual data, how
    would they predict a country's personal freedom school for one with a 6.75 rating
    for `pf_expression_control`? Is this an overestimate or an underestimate, and
    by how much? In other words, what is the residual for this prediction?

To predict, we will use the predict function, but we have to send the new data as a dataframe.

```{r}
predict(m1,newdata=data.frame(pf_expression_control=6.75))
```
We thus predict a value of 7.86 for the `pf_score`.

The observed value is 8.628272.

```{r}
hfi_sub %>%
  filter(pf_expression_control==6.75)
```
The residual is:

```{r}
8.628272 - 7.861402
```
We underestimated the actual value.

Another way to do this is to use the `broom` package.

```{r}
library(broom)
```

```{r}
augment(m1) %>%
   filter(pf_expression_control==6.75)
```


## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) 
linearity, (2) nearly normal residuals, and (3) constant variability.

**Linearity**: You already checked if the relationship between `pf_score` and 
`pf_expression_control' is linear using a scatterplot. We should also verify this
condition with a plot of the residuals vs. fitted (predicted) values.

```{r residuals}
ggplot(data = m1, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x="Fitted values",y="Residuals",title="Residual analysis") +
  theme_bw()
```

Notice here that `m1` can also serve as a data set because stored within it are 
the fitted values ($\hat{y}$) and the residuals. Also note that we're getting 
fancy with the code here. After creating the scatterplot on the first layer 
(first line of code), we overlay a horizontal dashed line at $y = 0$ (to help us
check whether residuals are distributed around 0), and we also rename the axis 
labels to be more informative and add a title.

>**Exercise**:  
Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between the two variables?  

The width is constant and there is no trend in the data. The linearity assumption is not bad.

**Nearly normal residuals**: To check this condition, we can look at a histogram

```{r hist-res}
ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(binwidth = .4) +
  xlab("Residuals") +
  theme_bw()
```

or a normal probability plot of the residuals.

```{r qq-res}
ggplot(data = m1, aes(sample = .resid)) +
  stat_qq() +
  theme_bw() +
  geom_abline(slope=1,intercept = 0)
```

Note that the syntax for making a normal probability plot is a bit different 
than what you're used to seeing: we set `sample` equal to the residuals instead 
of `x`, and we set a statistical method `qq`, which stands for "quantile-quantile",
another name commonly used for normal probability plots.

> **Exercise**: 
Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?

No, the sample is small but it appears the residual are skewed to the left.

**Constant variability**:

> **Exercise**: 
Based on the residuals vs. fitted plot, does the constant variability condition 
    appear to be met?

Yes, the width of the plot seems constant.

