---
title: "Multiple Regression Applications Solutions"
author:
- Lt Col Ken Horton
- Lt Col Kris Pruitt
- Professor Bradley Warner
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{multirow}
   - \usepackage{multicol}
   - \usepackage{changepage}
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align='center')
knitr::opts_chunk$set(out.width = "75%")
library(openintro)
library(knitr)
library(mosaic)
library(tidyverse)
```

\newcommand{\E}{\mbox{E}}
\newcommand{\Var}{\mbox{Var}}
\newcommand{\Cov}{\mbox{Cov}}
\newcommand{\Prob}{\mbox{P}}
\newcommand*\diff{\mathop{}\!\mathrm{d}}

\indent 1. The `mtcars` dataset contains average mileage (mpg) and other information about specific makes and models of cars. (This dataset is built-in to `R`; for more information about this dataset, reference the documentation with `?mtcars`). 

a. Build and interpret the coefficients of a model fitting `mpg` against displacement (`disp`), horsepower (`hp`), rear axle ratio (`drat`), and weight in 1000 lbs (`wt`). 

```{r}
cars_mod<-lm(mpg~disp+hp+drat+wt,data=mtcars)
summary(cars_mod)
```

$$
\E(\text{mpg})=29.15+0.004*\text{disp}-0.035*\text{hp}+1.768*\text{drat}-3.480*\text{wt}
$$

Each coefficient represents the expected increase in `mpg` for a unit increase in the respective variable, leaving all other variables constant. 

b) Given your model, what is the expected mpg for a vehicle with a displacement of 170, a horsepower of 100, a `drat` of 3.80 and a wt of 2,900 lbs. Construct a 95% confidence interval and prediction interval for that expected mpg. 

```{r}
predict(cars_mod,newdata=data.frame(disp=170,hp=100,drat=3.8,wt=2.9),interval="confidence")
predict(cars_mod,newdata=data.frame(disp=170,hp=100,drat=3.8,wt=2.9),interval="prediction")
```

c. Repeat part (b) with a bootstrap.

```{r}
set.seed(732)
results <- do(1000)*lm(mpg~disp+hp+drat+wt,data=resample(mtcars))
```

```{r}
head(results)
```


```{r}
results %>%
mutate(pred=Intercept+disp*170+hp*100+drat*3.8+wt*2.9) %>%
cdata(~pred,data=.)
```

```{r}
results %>%
mutate(pred=Intercept+disp*170+hp*100+drat*3.8+wt*2.9) %>%
cbind(resid=sample(cars_mod$residuals,size=1000,replace = TRUE)) %>%
mutate(pred_ind=pred+resid) %>%
cdata(~pred_ind,data=.)
```



&nbsp;

\indent 2. Is that the best model for predicting mpg? With your group, try a variety of different models. You could explore higher order terms or even interactions. One place to start is my using the `pairs()` function on `mtcars` to plot a large pairwise scatterplot. How high could you get adjusted $R$-squared? Keep in mind that is only one measure of fit. 

Answers will vary, but I tried this and got 0.8694. 

```{r}
summary(lm(mpg~disp+I(disp^2)+hp+I(hp^2)+wt,data=mtcars))
```




## File Creation Information 

  * File creation date: `r Sys.Date()`
  * Windows version: `r win.version()`
  * `r R.version.string`
  * `mosaic` package version: `r packageVersion("mosaic")`
  * `tidyverse` package version: `r packageVersion("tidyverse")`
  * `openintro` package version: `r packageVersion("openintro")`

 




