---
title: "Lesson 35 Application Solutions"
author: "Lt Col Ken Horton"
date: "November 18, 2019"
header-includes: 
  - \usepackage{amsmath,multirow}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align='center')
knitr::opts_chunk$set(out.width = "75%")
```


\newcommand{\E}{\mbox{E}}
\newcommand{\Var}{\mbox{Var}}
\newcommand{\Cov}{\mbox{Cov}}
\newcommand{\Prob}{\mbox{P}}
\newcommand*\diff{\mathop{}\!\mathrm{d}}

# Exercises

\indent 1. The `poisons.csv` dataset contains results of an experiment evaluating survival times of animals exposed to certain types of poisons and certain types of treatments.

a) Build a set of three boxplots that summarize survival time by type of poison. Describe the relationship you see. HINT: You may need to convert the `poison` variable to a factor prior to continuing. 
```{r lesson35a}
poison.data<-read.csv("poisons.csv")
poison.data$poison<-as.factor(poison.data$poison)
plot(poison.data$poison,poison.data$time)
```

Clearly, survival times are lower among type 3, but types 1 and 2 look relatively similar. 

b) Build a linear model fitting survival time against type of poison. Does there appear to be a significant difference in mean survival time by type of poison? 

```{r lesson35b}
anova(lm(time~poison,data=poison.data))
```

Yes, there appears to be a difference in mean survival time based on type of poison. 

c) Write out the estimated linear model. In words, interpret the coefficient estimates. 
```{r lesson35d}
summary(lm(time~poison,data=poison.data))
```

$$
Y=0.6175 - 0.0731x_1 - 0.3413x_2
$$

The estimate of $\beta_0$, 0.6175, represents the mean survival time among the reference category, poison type 1. The estimate of $\beta_1$, -0.073, represents the mean difference in survival time between type 2 and type 1. Similarly, the estimate of $\beta_2$, -0.341, represents the mean difference in survival time between type 3 and type 1. 

d) Obtain pairwise differences to determine where the significant differences lie. 
```{r lesson35c}
TukeyHSD(aov(time~poison,data=poison.data))
```

Types 1 and 2 don't appear to differ, but type 3 seems to be result in significantly lower mean survival time than types 1 and 2. 


